---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggcapdthemes

<!-- badges: start -->
<!-- badges: end -->

The `ggcapdthemes` package provides a custom ggplot2 theme and functionality to incorporate guidance from the EPA's Clean Air Power Division (CAPD) Data Visualization Style Guide. Examples show how to implement the theme and color palettes and demonstrate best practices for building various chart types in R.

## Installation

You can install the development version of ggcapdthemes from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
#NOTE: You will need the “devtools” package to utilize the install_github function
# install.packages("devtools")
devtools::install_github("usepa/ggcapdthemes", dependencies = TRUE, build_vignettes = TRUE)
```

## Example

Here we use the `resource_mix` dataset from eGRID 2018-2023 Summary data. We show the barplot using ggplot2 and and after applying the `scale_fill_capd_discrete` and `theme_capd` functions.

```{r example}
library(ggplot2)
library(ggcapdthemes)

## original chart
gg_ex_bar

## with capd theme
gg_ex_bar + theme_capd(plot_direction = 'vert')

## with fuel type color palette and capd theme
gg_ex_bar +
  scale_fill_capd_discrete(palette = 'fuel_type') +
  theme_capd(plot_direction  = 'vert', legend_position = 'none') 
```

## Disclaimer

The United States Environmental Protection Agency (EPA) GitHub project code is provided on an "as is" basis and the user assumes responsibility for its use.  EPA has relinquished control of the information and no longer has responsibility to protect the integrity , confidentiality, or availability of the information.  Any reference to specific commercial products, processes, or services by service mark, trademark, manufacturer, or otherwise, does not constitute or imply their endorsement, recommendation or favoring by EPA.  The EPA seal and logo shall not be used in any manner to imply endorsement of any commercial product or activity by EPA or the United States Government.
