---
title: "Using ggcapdthemes and `theme_capd`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Using ggcapdthemes and `theme_capd`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 8, fig.height=4, dev='svg'
)
```

## Introduction

Welcome! This vignette will demonstrate how the EPA CAPD Data Visualization Style Guide can be integrated into your R visualizations. Through the use of the `{ggcapdthemes}` package and the `theme_capd` function, you can easily customize and standardize your ggplots!

```{r, echo = FALSE}
knitr::include_graphics('chart-layout.svg')
```

Because this package is intended to augment the functionality of ggplot2, we suggest loading it prior to loading `ggcapdthemes`.

```{r}
library(ggplot2)
library(ggcapdthemes)
```

For R users looking for DVSG best practices that go beyond the theme, check out the [Best Practices](best_practices.html) vignette.

## Applying the theme: `theme_capd()`

First, we start with some eGRID data shown in a ggplot bar chart, using the default ggplot theme. The `ggcapdthemes` package contains a data object called `gg_ex_bar` to use as a building block.

```{r}

gg_ex_bar

```

Next, we add the `theme_capd()` function to apply our custom theme.

```{r}
gg_ex_bar +
  theme_capd()
```

To change the orientation of the axis and grid lines to the y-axis, we can provide the `axis_lines = 'vert'` argument. The default is 'horiz' for horizontal (x-axis and horizontal grid lines shown).

```{r}
gg_ex_bar +
  theme_capd(axis_lines = 'vert')
```

Other theme specifications can be controlled with the `ticks`, `grid_lines`, and `legend_position` arguments, as shown here.

```{r}
gg_ex_bar +
  theme_capd(grid_lines = c('major','minor'), 
             axis_lines = 'vert', 
             ticks = TRUE, 
             legend_position = 'none')
```

Lastly, we can add the `scale_fill_capd_discrete` function to apply a custom color palette for fuel types.

```{r}
gg_ex_bar +
  scale_fill_capd_discrete(palette = 'fuel_type') +
  theme_capd(grid_lines = c('major','minor'), 
             axis_lines = 'vert', 
             ticks = TRUE, 
             legend_position = 'none')

```

### General theme info

The "Source Sans 3" font is downloaded from Google Fonts and applied to all when `theme_capd` is used. An initial font size is specified using the `base_size` argument; this defaults to 11pt, but we will use 10pt for this example. 

### Font sizing: `base_size`

Other fonts throughout the chart are re-sized relative to the `base_size`, using the `ggplot2::rel` function. The `half_line` setting is used for margins.

```{r}
gg_ex_bar +
  theme_capd(base_size = 11)

gg_ex_bar +
  theme_capd(base_size = 6)

gg_ex_bar +
  theme_capd(base_size = 16)
```

Here we see the font applied to all text elements by default; they will be re-sized in the sections below.

Plot annotations such as titles, subtitles is formatted through the following theme options. The title size is magnified 20% compared to `base_size`, while the caption is decreased `20%`.

<a href="#TOC">Return to Table of Contents</a>

### Axis line orientation: `axis_lines`

```{r}
## default: 'horiz'
gg_ex_bar +
  theme_capd(axis_lines = 'horiz')

gg_ex_bar +
  theme_capd(axis_lines = 'vert')

gg_ex_bar +
  theme_capd(axis_lines = 'both')
```

<a href="#TOC">Return to Table of Contents</a>

### Gridlines: `grid_lines`

The `grid_lines` arguments controls which sets of gridlines to display. 

Possible options include

- major grid lines `'major'` 
- minor grid lines: `'minor'` 
- both major and minor: `c('major','minor')`
- no gridlines: `'none'`


Note that this interacts with the `axis_lines` argument and only shows the gridlines for the orientations specified. Other gridlines remain hidden.

```{r}
gg_ex_bar +
  theme_capd(grid_lines = 'major', axis_lines = 'horiz')

gg_ex_bar +
  theme_capd(grid_lines = 'major', axis_lines = 'vert')

gg_ex_bar +
  theme_capd(grid_lines = c('major','minor'), axis_lines = 'vert')

gg_ex_bar +
  theme_capd(grid_lines = 'none')
```

<a href="#TOC">Return to Table of Contents</a>

### Tick Marks: `ticks` 

The `ticks` argument determines whether to show (TRUE) or hide (FALSE) the tick marks on the non-major axis, or the one perpendicular to the plot's orientation. For vertical `axis_liness`, this is the y-axis ticks marks and for horizontal orientations, this is the x-axis tick marks.

```{r}
gg_ex_bar +
  theme_capd(ticks = FALSE)

gg_ex_bar +
  theme_capd(ticks = TRUE)
```

When `axis_lines = 'both'`, this controls whether to show both x-axis and y-axis tick marks.

```{r}
gg_ex_bar +
  theme_capd(ticks = TRUE, axis_lines = 'both')
```

<a href="#TOC">Return to Table of Contents</a>

### Legend: `legend_position`

This is a wrapper for the `legend.position` ggplot theme argument. Valid options for `legend_position` are 'top', 'bottom', 'left', 'right', or 'none' to hide it.

```{r}
gg_ex_bar +
  theme_capd(legend_position = 'top')

gg_ex_bar +
  theme_capd(legend_position = 'bottom')

gg_ex_bar +
  theme_capd(legend_position = 'none')
```

<a href="#TOC">Return to Table of Contents</a>

## Applying color palettes

Color scales are added separately from the theme, using the `scale_color_capd_*` and `scale_fill_capd_*` functions. Here we apply the "unordered" palette using `scale_color_capd_discrete`.

### Quantitative palettes

#### Sequential

```{r, echo=F}
op=par(mar=c(0,2,0,1))
pals::pal.bands(
  ggcapdthemes:::capd_pal_d('sequential')(7),
  ggcapdthemes:::capd_pal_d('sequential')(6),
  ggcapdthemes:::capd_pal_d('sequential')(5),
  ggcapdthemes:::capd_pal_d('sequential')(4),
  ggcapdthemes:::capd_pal_d('sequential')(3),
  labels=c('n = 7','n = 6','n = 5', 'n = 4', 'n = 3')
)
par(op)
```

#### Divergent

```{r, echo=F}
op=par(mar=c(0,2,0,1))
pals::pal.bands(
  ggcapdthemes:::capd_pal_d('divergent')(7),
  ggcapdthemes:::capd_pal_d('divergent')(6),
  ggcapdthemes:::capd_pal_d('divergent')(5),
  ggcapdthemes:::capd_pal_d('divergent')(4),
  ggcapdthemes:::capd_pal_d('divergent')(3),
  labels=c('n = 7','n = 6','n = 5', 'n = 4', 'n = 3')
)
par(op)
```

```{r,echo=F}
op=par(mar=c(0,2,0,1))
pals::pal.bands(
  ggcapdthemes:::capd_pal_d('divergent2')(7),
  ggcapdthemes:::capd_pal_d('divergent2')(6),
  ggcapdthemes:::capd_pal_d('divergent2')(5),
  ggcapdthemes:::capd_pal_d('divergent2')(4),
  ggcapdthemes:::capd_pal_d('divergent2')(3),
  labels=c('n = 7','n = 6','n = 5', 'n = 4', 'n = 3')
)
par(op)
```

### Categorical palettes

#### Fuel type palettes

```{r, eval=T,echo=F}
op=par(mar=c(0,2,0,1))
pals::pal.bands(capd_palettes$fuel_type,
                capd_palettes$fuel_type2,
                labels = c('fuel_type', 'fuel_type2'),
                show.names = T)
par(op)
```

Example of a barchart using the `scale_fill_capd_discrete` function

```{r}
ggplot(resource_mix[resource_mix$year==2023,],aes(y = resource_label, x = generation_resource_mix, fill = resource_label)) +
    geom_bar(stat='identity') +
    scale_fill_capd_discrete(palette='fuel_type')+
    scale_x_continuous(expand=c(0,0), labels=scales::label_percent())+
    labs(x='', y='Resource Mix (%)', title='eGRID 2023 Resource Mix', color='Resource') +
    theme_capd(axis_lines = 'vert', grid_lines = c('major','minor'), legend_position = 'none') 
```

Example of a line plot using the `scale_color_capd_discrete` function

```{r}
ggplot(resource_mix,aes(x = year, y = generation_resource_mix*100, color = resource_label, group = resource_label)) +
  geom_line() +
  geom_point()+
  scale_color_capd_discrete(palette='fuel_type')+
  labs(x='', y='Resource Mix (%)', title='eGRID Annual Resource Mix, 2018-2023', color='Resource') +
  theme_capd(grid_lines = c('major','minor'))
```

##### Combustion vs Non-Combustion

All Combustion fuel types:

Includes all fossil fuels (coal, oil, natural gas, blast furnace gas, and tire-derived fuel) and biomass, such as wood, agricultural byproducts, municipal solid waste, and landfill gas.

```{r}
ggplot(combustion) +
  geom_bar(aes(x = resource_mix, y=region, group=region, fill=fuel_type), stat='identity') +
  scale_fill_capd_discrete(palette = 'combustion') +
  scale_x_continuous(expand = c(0,0)) + 
  labs(x = 'Resource Mix (%)', y='eGRID Subregion', fill = 'Fuel Type', title = '2023 eGRID Resource Mix',
       subtitle = 'by Combustion vs. Non-combustion fuels') +
  theme_capd(axis_lines = 'vert')
```

##### Renewables vs Non-renewables

```{r, echo=F}
tmp <- capd_palettes$renewables
names(tmp)[1:2] <- c("Non-renewables\n(excluding nuclear)",
                "Renewables\n(excluding hydro)")
op=par(mar=c(0,3,0,1))
pals::pal.bands(tmp, capd_palettes$renewables2,
                labels = c('renewables', 'renewables2'),
                show.names = T)
par(op)
```

```{r}
ggplot(renewables) +
  geom_bar(aes(x = resource_mix, y = region, group = region, fill = fuel_type), stat = 'identity') +
  scale_fill_capd_discrete(palette = 'renewables2') +
  scale_x_continuous(expand=c(0,0)) +
  labs(x = 'Resource Mix (%)', y='eGRID Subregion', fill = 'Fuel Type', title = '2023 eGRID Resource Mix',
       subtitle = 'by Renewable vs. Non-renewable fuels') +
  theme_capd(axis_lines = 'vert')
```

#### Pollutant palettes

```{r, echo=F}
op=par(mar=c(0,2,0,1))
pals::pal.bands(capd_palettes$pollutant)
par(op)
```

#### Unordered palettes

```{r, echo=F, fig.height=4}
op=par(mar=c(0,2,0,1))
pals::pal.bands(capd_palettes$unordered,
                capd_palettes$unordered2,
                labels = c('unordered','unordered2'),
                show.names = T)
par(op)
```
